Concurrent Programming with CXL

### 2020-06-29 book/paper.tex
Formatting up to Chapter 18

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \textit{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}%
There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%
If \texttt{c} is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

The CXL \texttt{synch} module supports Mesa condition variables.
However, like C, Java, and Python,
it does not have built-in language support for them but instead associates condition variables
with explicit locks.
Figure~\ref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition(lk)} creates a new condition variable associated with the given pointer
to a lock \texttt{lk}.
The condition variable itself is represented by a dictionary containing the pointer to the lock
and a bag of name tags of processes waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

In \texttt{release\_rlock}, notice that \texttt{notify(\&wcond)} is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll(\&wcond)} as well as
\texttt{notify(\&wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.

>>>>

Tony Hoare, who came up with the concept of split binary semaphores, devised
an abstraction of the concept in a programming language paradigm called
\emph{monitors}~\cite{Hoare74}.
\index{monitor}%
(A similar construct was independently invented by Per Brinch Hansen~\cite{BH73}.)
A monitor is a special version of an object-oriented \emph{class}, comprising
a set of variables and methods that operate on those variables.
There is a split binary semaphore associated with each such class.
The \textit{mutex} is hidden: it is automatically acquired when invoking a
method and released upon exit.
The other semaphores are called \emph{condition variables}.
\index{condition variable}%
There are two operations on condition variables: \texttt{wait}
\index{wait}%
and
\texttt{signal}.
\index{signal}%
If $c$ is a pointer to a condition variable, then the semantics of the
operations, in CXL, are as follows:

@@

The CXL \texttt{synch} module supports Mesa condition variables.
However, like C, Java, and Python,
it does not have built-in language support for them but instead associates condition variables
with explicit locks.
Figure~\ref{fig:cv} shows the implementation of
condition variables in the \texttt{synch} module.
\texttt{Condition}(\textit{lk}) creates a new condition variable associated with the given pointer
to a lock \textit{lk}.
The condition variable itself is represented by a dictionary containing the pointer to the lock
and a bag of name tags of processes waiting on the condition variable.
(The \texttt{synchS} library instead uses a list of contexts.)

@@

In \texttt{release\_rlock}, notice that \texttt{notify}(\&\textit{wcond}) is invoked when there
are no readers left, \emph{without} checking if there are writers waiting to enter.
With Mesa monitors this is ok, because calling \texttt{notify} is a no-op if no process
is waiting.

\texttt{release\_wlock} executes \texttt{notifyAll}(\&\textit{wcond}) as well as
\texttt{notify}(\&\textit{wcond)}.
Again, because we do not keep track of the number of waiting readers or writers, we
have to conservatively assume that all waiting readers can enter, or, alternatively,
up to one waiting writer can enter.  So \texttt{release\_wlock} wakes up all
potential candidates.
There are two things to note here.  First, unlike split binary semaphores or Hoare
monitors, where multiple waiting readers would have to be signaled one at a time in a
baton-passing fashion (see Figure~\ref{fig:RWsplitsema}), with Mesa monitors
all readers are awaked in one fell swoop using \texttt{notifyAll}.
Second, both readers and writers are awakened---this is ok because both execute
\texttt{wait} within a \textbf{while} loop, re-checking the condition that they
are waiting for.  So if both type of processes are waiting, either all the readers
get to enter next or one of the writers gets to enter next.
